---
title: "Week 4: Chart Types & Variations"
output:
  html_document:
    toc: true
    toc_float: true
---

<div class="hero-banner">
<div class="week-badge">4</div>
<h1>Chart Types &amp; Variations</h1>
<p>Beyond bar charts &mdash; a field guide to the visualization zoo</p>
</div>

::: {.objectives}
## Learning Objectives

By the end of this week, you will be able to:

- Choose the right chart type for a given data relationship
- Create lollipop charts, dumbbell charts, ridgeline plots, treemaps, bubble charts, heatmaps, and correlation matrices
- Understand when to use each chart type and articulate why one choice is better than another
- Apply the principle of matching visual encoding to the analytical question
:::

## Setup

Before we begin, load the core packages we will use throughout this week. Several chart types require specialized packages that extend ggplot2. If you have not installed them yet, see the Common Errors box below.

```{r setup-week04, message=FALSE, warning=FALSE}
library(tidyverse)
library(ggridges)
library(treemapify)
library(gapminder)
library(GGally)
```

::: {.callout-error}
**Common Errors You May Encounter This Week**

This week uses several extension packages. If you run into errors, check this list first:

1. **`Error: could not find function "geom_density_ridges_gradient"`**
   This function lives in the `ggridges` package, which is not part of the tidyverse. You need to install it once and then load it:
   ```r
   install.packages("ggridges")
   library(ggridges)
   ```

2. **`Error: could not find function "geom_treemap"`**
   This function lives in the `treemapify` package. Install and load it:
   ```r
   install.packages("treemapify")
   library(treemapify)
   ```

3. **`ggpairs()` takes forever to run / R appears frozen**
   The `ggpairs()` function computes pairwise plots for every combination of columns. If you pass a dataset with many columns, the computation time grows quadratically. Always use `select()` to pick only 4--5 numeric columns before passing to `ggpairs()`:
   ```r
   # Good -- 5 columns = 10 pairs
   mtcars %>% select(mpg, disp, hp, wt, qsec) %>% ggpairs()

   # Bad -- 11 columns = 55 pairs (slow!)
   mtcars %>% ggpairs()
   ```

4. **Bubble chart sizes look wrong / small countries invisible**
   Use `scale_size_area()` instead of `scale_size_continuous()`. The `scale_size_area()` function ensures that the *area* of each bubble is proportional to the data value, which is the perceptually correct mapping. With `scale_size_continuous()`, the *radius* is mapped linearly, causing larger values to appear disproportionately huge.

5. **Heatmap text is unreadable / labels disappear into dark tiles**
   Adjust the `size` parameter in `geom_text()` to make labels larger or smaller. On dark tiles, switch the text `color` to `"white"` or a light value. You can also use conditional coloring:
   ```r
   geom_text(aes(label = round(value, 1),
                 color = after_stat(value > threshold)), size = 3)
   ```
   Or simply pick a text color that contrasts well with your palette (e.g., `"white"` on dark sequential scales).
:::

## Choosing the Right Chart

The single most important decision in data visualization is choosing the right chart type. A beautiful chart that uses the wrong encoding is worse than a plain chart that uses the right one. The chart type you select should be driven entirely by the **relationship** you want to show in your data.

Here is a framework for thinking about chart selection:

- **Comparison** of values across categories: bar chart, lollipop chart, dumbbell chart, dot plot
- **Distribution** of a single variable or across groups: histogram, density plot, violin plot, ridgeline plot, boxplot
- **Relationship** between two or more variables: scatterplot, bubble chart, heatmap, correlogram
- **Composition** showing parts of a whole: stacked bar chart, treemap, waffle chart
- **Change over time**: line chart, area chart, sparkline

::: {.callout-info}
**There is no universal "best" chart.** The best chart is the one that most clearly communicates the specific insight you want your audience to take away. A bar chart is not inherently better or worse than a treemap -- it depends on the question you are answering.
:::

Each of the following sections introduces a chart type, explains when it is most useful, and provides working R code you can adapt for your own projects.

## Try It: Chart Type Explorer

You have just seen that different chart types serve different analytical purposes. Now explore them interactively. The sandbox below lets you switch between chart types using the same dataset, so you can see how the choice of chart changes the story.

<div class="sandbox-section">
<div class="sandbox-header">
<span class="sandbox-icon">🧪</span> Chart Type Explorer — Pick the Right Tool for the Job
</div>
<iframe class="sandbox-frame" src="https://analytics.gonzaga.edu/content/84662c36-158f-47a8-ae95-5f9f7b0e4652/"
        height="600" loading="lazy" allowfullscreen></iframe>
<div class="sandbox-cold-start">
If the app takes a few seconds to load on first visit, that is normal — the server is waking up.
</div>
</div>

::: {.try-it-box}
**Exploration Tasks:**

1. Start with the **bar chart** view. What comparison does it emphasize?
2. Switch to a **lollipop chart** — how does it compare to the bar chart for readability?
3. Try the **heatmap** — what relationships does it reveal that the other charts do not?
4. Look at the **treemap** — when might this be more useful than a bar chart?
:::

::: {.callout-info}
**What You Should Have Noticed:** No single chart type is "best." Each one highlights different aspects of the data. Bar charts excel at precise comparisons, lollipop charts reduce visual clutter, heatmaps reveal patterns across two dimensions, and treemaps show part-to-whole relationships with hierarchy. The right choice depends on what question you are trying to answer.
:::

::: {.callout-ai}
**AI & This Concept**
When asking AI to create a visualization, always tell it *what comparison you want to make*, not just "chart this data." For example, "Show how categories rank from highest to lowest" will get you a sorted bar chart, while "Show correlations between all numeric variables" will get a heatmap. The more specific your intent, the better the chart.
:::

## Lollipop Charts {.tabset}

Lollipop charts are an elegant alternative to bar charts. They convey the same information -- a categorical variable mapped to a quantitative value -- but with far less ink. Edward Tufte would approve: the data-to-ink ratio is much higher.

Use a lollipop chart when you have a **ranked comparison** of values across categories, especially when you have many categories where thick bars would feel heavy.

### The Code

```{r lollipop-chart, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
mpg %>%
  group_by(manufacturer) %>%
  summarise(avg_hwy = mean(hwy)) %>%
  mutate(manufacturer = fct_reorder(manufacturer, avg_hwy)) %>%
  ggplot(aes(x = avg_hwy, y = manufacturer)) +
  geom_segment(aes(x = 0, xend = avg_hwy, yend = manufacturer),
               color = "#94a3b8", linewidth = 0.8) +
  geom_point(color = "#002967", size = 3) +
  labs(title = "Average Highway MPG by Manufacturer",
       subtitle = "Lollipop chart — less ink than a bar chart",
       x = "Average Highway MPG", y = NULL) +
  theme_minimal(base_size = 13)
```

### How It Works

The lollipop chart is built from two geoms layered together:

1. `geom_segment()` draws the thin line (the "stick") from zero to the data value
2. `geom_point()` draws the dot (the "candy") at the end

Notice that we use `fct_reorder()` to sort the manufacturers by their average MPG. This is critical -- an unsorted lollipop chart is much harder to read. Always sort by the quantitative variable unless there is a natural ordering (like months or rankings) that takes priority.

::: {.callout-tip}
**When to choose a lollipop over a bar chart:** Lollipop charts shine when you have 8 or more categories. With fewer categories, a standard bar chart is fine. With many categories, the thin stems of a lollipop chart reduce visual clutter significantly.
:::

## Dumbbell Charts

Dumbbell charts (also called DNA charts or connected dot plots) show the **difference between two values** for each category. They are ideal for before/after comparisons, paired measurements, or showing the gap between two groups.

The "dumbbell" shape -- two dots connected by a line -- naturally draws the eye to the distance between the two values, making it easy to compare gaps across categories.

```{r dumbbell-chart, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
dumbbell_data <- mpg %>%
  group_by(manufacturer) %>%
  summarise(city = mean(cty), highway = mean(hwy)) %>%
  mutate(manufacturer = fct_reorder(manufacturer, highway))

ggplot(dumbbell_data) +
  geom_segment(aes(x = city, xend = highway, y = manufacturer, yend = manufacturer),
               color = "#94a3b8", linewidth = 1.2) +
  geom_point(aes(x = city, y = manufacturer), color = "#C41E3A", size = 3) +
  geom_point(aes(x = highway, y = manufacturer), color = "#002967", size = 3) +
  labs(title = "City vs. Highway MPG by Manufacturer",
       subtitle = "Red = City | Navy = Highway",
       x = "Miles Per Gallon", y = NULL) +
  theme_minimal(base_size = 13)
```

The segment connects the two values, and the colored dots mark each endpoint. The color legend is embedded in the subtitle to keep the chart clean. Notice how easy it is to see which manufacturers have the largest gap between city and highway fuel economy.

::: {.callout-tip}
**Dumbbell charts are underused.** Many analysts default to grouped bar charts for paired comparisons, but dumbbell charts are almost always more readable. The connecting line makes the comparison explicit, while grouped bars force the viewer to mentally bridge the gap.
:::

## Ridgeline Plots (Joy Plots)

Ridgeline plots display the **distribution** of a variable across multiple groups, stacked vertically with slight overlap. They are one of the most visually striking chart types in the ggplot2 ecosystem, and they are remarkably effective at revealing distributional differences.

The name "joy plot" comes from the iconic cover of Joy Division's *Unknown Pleasures* album, which featured a similar stacked-ridge visualization of radio pulsar data.

```{r ridgeline-plot, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
ggplot(mpg, aes(x = hwy, y = fct_reorder(class, hwy, .fun = median), fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = 2, rel_min_height = 0.01) +
  scale_fill_viridis_c(option = "plasma") +
  labs(title = "Distribution of Highway MPG by Vehicle Class",
       subtitle = "Ridgeline plot — compact display of many distributions",
       x = "Highway MPG", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

The `scale` parameter controls how much the ridges overlap. A value of 2 means each ridge can extend up to twice the allocated row height. The `rel_min_height` parameter trims the tails to avoid distracting wisps at the edges.

The gradient fill using `after_stat(x)` maps the x-axis value to a color, reinforcing the positional encoding with a color encoding. This is a form of **redundant encoding** -- using two visual channels for the same variable -- which can improve readability.

::: {.callout-info}
**When to use ridgeline plots:** They work best when you have 5--15 groups and want to compare the full shape of each distribution. With fewer groups, consider overlapping density plots or faceted histograms. With more groups, the overlap becomes too dense to read.
:::

## Treemaps

Treemaps display **hierarchical or part-to-whole** relationships using nested rectangles. The area of each rectangle is proportional to a quantitative variable, making treemaps an effective way to show composition and relative size simultaneously.

Treemaps are particularly useful when you have two levels of hierarchy (e.g., manufacturer and class, or department and team) and want to show how subcategories contribute to larger groups.

```{r treemap-chart, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
mpg %>%
  count(manufacturer, class) %>%
  ggplot(aes(area = n, fill = manufacturer, subgroup = manufacturer, label = class)) +
  geom_treemap() +
  geom_treemap_text(color = "white", place = "centre", size = 10) +
  geom_treemap_subgroup_border(color = "white", size = 3) +
  labs(title = "Vehicle Models: Manufacturer x Class",
       subtitle = "Area represents number of models in the mpg dataset") +
  theme(legend.position = "none")
```

The `treemapify` package provides ggplot2-compatible geoms for treemaps. `geom_treemap()` draws the filled rectangles, `geom_treemap_text()` adds labels inside each rectangle, and `geom_treemap_subgroup_border()` draws thicker borders around the top-level grouping (manufacturer) to make the hierarchy visible.

::: {.callout-tip}
**Treemaps vs. pie charts:** Both show composition, but treemaps are far more effective because we can judge rectangular areas more accurately than circular arc lengths. Treemaps also scale to many more categories and support nested hierarchies.
:::

## Bubble Charts

A bubble chart is a scatterplot where a **third variable** controls the size of each point. When done well, bubble charts can pack an impressive amount of information into a single graphic. The most famous example is Hans Rosling's Gapminder visualization of global health and wealth.

```{r bubble-chart-gapminder, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
gapminder %>%
  filter(year == 2007) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp, size = pop, color = continent)) +
  geom_point(alpha = 0.6) +
  scale_x_log10(labels = scales::dollar_format()) +
  scale_size_area(max_size = 15, guide = "none") +
  scale_color_manual(values = c("Africa" = "#C41E3A", "Americas" = "#002967",
                                "Asia" = "#B4975A", "Europe" = "#2563eb", "Oceania" = "#10b981")) +
  labs(title = "Wealth, Health, and Population (2007)",
       subtitle = "Bubble size = population",
       x = "GDP per Capita (log scale)", y = "Life Expectancy",
       color = "Continent",
       caption = "Source: Gapminder Foundation") +
  theme_minimal(base_size = 13)
```

Several design decisions make this chart work:

- **Log scale on x-axis**: GDP per capita has an extremely skewed distribution. A log scale spreads the data out so that both poor and rich countries are visible.
- **`scale_size_area()`**: This ensures that the area (not the radius) of each bubble is proportional to the population, which is the perceptually correct mapping.
- **Alpha transparency**: With many overlapping bubbles, transparency prevents large countries from completely obscuring small ones.
- **Color by continent**: This adds a fourth variable, creating a chart that encodes GDP, life expectancy, population, and continent in a single view.

::: {.callout-info}
**Caution with bubble charts:** Do not use more than one size variable. Human perception of area is imprecise, so bubble charts work best when the size variable is used for rough comparisons ("China is much bigger than Norway") rather than precise readings. If you need precision, use a different encoding.
:::

## Heatmaps

Heatmaps use **color intensity** to represent values in a matrix. They are excellent for spotting patterns, clusters, and outliers in data that has two categorical dimensions and one quantitative dimension.

```{r heatmap-chart, fig.width=8, fig.height=6, message=FALSE, warning=FALSE}
mpg %>%
  group_by(manufacturer, class) %>%
  summarise(avg_hwy = mean(hwy), .groups = "drop") %>%
  ggplot(aes(x = class, y = manufacturer, fill = avg_hwy)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = round(avg_hwy, 0)), size = 3, color = "white") +
  scale_fill_viridis_c(option = "mako", direction = -1, na.value = "#f1f5f9") +
  labs(title = "Average Highway MPG: Manufacturer x Class",
       fill = "Avg MPG", x = NULL, y = NULL) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

Key elements of a good heatmap:

- **White borders** between tiles create visual separation and prevent colors from blending together.
- **Text labels** inside tiles let readers extract exact values, compensating for the imprecision of color perception.
- **A sequential color scale** (like viridis "mako") maps low values to light colors and high values to dark (or vice versa). Avoid rainbow color scales, which create false perceptual boundaries.
- **Missing data handling** with `na.value` ensures that empty cells are visually distinct from low-value cells.

The pattern of missing tiles (where no manufacturer makes a vehicle in that class) is itself informative -- it reveals the market segmentation strategy of each brand.

## Correlation Matrices

A correlation matrix visualizes the **pairwise relationships** between all numeric variables in a dataset. The `GGally` package provides `ggpairs()`, which creates a matrix of scatterplots, correlation coefficients, and density curves in a single call.

```{r correlation-matrix-ggally, fig.width=7, fig.height=7, message=FALSE, warning=FALSE}
mtcars %>%
  select(mpg, disp, hp, wt, qsec) %>%
  ggpairs(
    upper = list(continuous = wrap("cor", color = "#002967")),
    lower = list(continuous = wrap("smooth", alpha = 0.3, color = "#002967")),
    diag = list(continuous = wrap("densityDiag", fill = "#eff6ff", color = "#002967"))
  ) +
  labs(title = "Correlation Matrix — mtcars") +
  theme_minimal(base_size = 11)
```

The matrix has three regions:

- **Upper triangle**: Correlation coefficients (numeric summaries)
- **Lower triangle**: Scatterplots with smoothed trend lines (visual relationships)
- **Diagonal**: Density plots (individual distributions)

This layout is efficient because the upper and lower triangles are mirror images -- they show the same pairs, just in different forms. Together, they give you both the number and the shape of each relationship.

::: {.callout-tip}
**Reading correlations:** Look for the strongest relationships first. In the mtcars data, `wt` and `mpg` have a strong negative correlation (heavier cars get worse mileage), while `disp` and `hp` have a strong positive correlation (bigger engines tend to be more powerful). These are exactly the patterns you would expect from domain knowledge, which is reassuring.
:::

## Violin Plots

Violin plots combine the summary statistics of a boxplot with the distributional shape of a density plot. The "violin" shape shows where the data is concentrated, while the embedded boxplot marks the median and interquartile range.

```{r violin-plot-class, fig.width=8, fig.height=5, message=FALSE, warning=FALSE}
ggplot(mpg, aes(x = fct_reorder(class, hwy, .fun = median), y = hwy, fill = class)) +
  geom_violin(alpha = 0.6, color = NA) +
  geom_boxplot(width = 0.15, fill = "white", outlier.size = 1) +
  scale_fill_brewer(palette = "Set3") +
  labs(title = "Highway MPG by Vehicle Class",
       subtitle = "Violin + boxplot overlay",
       x = NULL, y = "Highway MPG") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

The combination of violin and boxplot is more informative than either alone:

- The **violin** shows whether the distribution is unimodal, bimodal, or skewed -- information that a boxplot hides entirely.
- The **boxplot** adds precise reference points (median, quartiles) that are difficult to read from a density shape alone.

Notice that we set `color = NA` on the violin to remove the outline and `fill = "white"` on the boxplot so it sits cleanly inside the violin shape.

## A Decision Framework

When you are faced with a new dataset and need to choose a chart type, work through these questions:

**"What is my primary question?"**

| I want to...                    | Consider these chart types                    |
|:--------------------------------|:----------------------------------------------|
| Compare values across groups    | Bar chart, lollipop chart, dumbbell chart     |
| Show a distribution             | Histogram, density plot, violin, ridgeline    |
| Show a relationship             | Scatterplot, bubble chart, heatmap            |
| Show composition / parts-whole  | Stacked bar chart, treemap, waffle chart      |
| Show change over time           | Line chart, area chart, sparkline             |

**"How many categories do I have?"**

- 2--5 categories: Almost any chart type works
- 6--15 categories: Lollipop, ridgeline, and heatmaps handle this range well
- 15+ categories: Consider filtering, grouping, or using small multiples

**"Who is my audience?"**

- General public: Stick to familiar types (bar, line, scatter)
- Technical audience: You can use more specialized types (ridgeline, correlogram, treemap)
- Decision makers: Prioritize clarity and annotate key takeaways directly on the chart

::: {.callout-info}
**The "one more chart" rule:** If you have time, always try at least one alternative chart type for your data. You may discover that a different encoding reveals a pattern you missed. Visualization is an iterative process.
:::

## Ignatian Reflection

::: {.callout-ignatian}
**"Men and Women for Others" -- Chart Choice as an Act of Service**

Choosing the right chart type is not merely a technical skill -- it is an act of service to your audience. When you select a visualization that clearly communicates a pattern, you empower others to understand and act. When you choose a chart that obscures or misleads, you fail in that responsibility.

St. Ignatius spoke of *Cura Personalis* -- care for the whole person. In the context of data visualization, this means considering your audience's background, literacy, and needs. A ridgeline plot may be beautiful and information-rich, but if your audience has never seen one before, a simpler chart with careful annotation may serve them better. The goal is not to showcase your technical skills but to illuminate truth.

The wrong chart can mislead; the right one illuminates. As you practice building different chart types this week, ask yourself: *Am I choosing this chart because it serves my audience, or because it looks impressive?* The answer to that question is the difference between decoration and communication.
:::

## Exercises

::: {.exercise-block}
::: {.exercise-header}
Week 4 Exercises
:::
::: {.exercise-body}

**Exercise 1: Lollipop Chart with Gapminder**

Using the `gapminder` package, filter the data to the year 2007 and a single continent of your choice. Create a lollipop chart showing life expectancy by country, sorted from highest to lowest. Use Gonzaga navy (`#002967`) for the dots and a light gray for the stems. Add a meaningful title and subtitle.

Fill in the blanks in the starter code below:

```{r exercise-lollipop-gapminder, eval=FALSE}
library(gapminder)

gapminder %>%
  filter(year == 2007, continent == "___") %>%
  mutate(country = fct_reorder(country, lifeExp)) %>%
  ggplot(aes(x = lifeExp, y = country)) +
  geom_segment(aes(x = 0, xend = ___, yend = ___),
               color = "#94a3b8", linewidth = 0.8) +
  geom_point(color = "___", size = 3) +
  labs(title = "___",
       subtitle = "___",
       x = "Life Expectancy (years)", y = NULL) +
  theme_minimal(base_size = 12)
```

**Hints:**

- The `xend` in `geom_segment()` should be the same variable mapped to the x-axis.
- The `yend` should be the same variable mapped to the y-axis.
- Gonzaga navy is `#002967`.

---

**Exercise 2: Ridgeline Plot of Diamond Prices**

Using the built-in `diamonds` dataset, create a ridgeline plot showing the distribution of `price` by `cut` quality. Use `fct_reorder()` to sort the cuts by median price. Experiment with the `scale` parameter and choose a color palette that works well.

Fill in the blanks:

```{r exercise-ridgeline-diamonds, eval=FALSE}
library(ggridges)

ggplot(diamonds, aes(x = ___, y = fct_reorder(___, ___, .fun = median),
                     fill = after_stat(x))) +
  geom_density_ridges_gradient(scale = ___, rel_min_height = 0.01) +
  scale_fill_viridis_c(option = "___") +
  labs(title = "___",
       subtitle = "___",
       x = "Price (USD)", y = NULL) +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none")
```

**Hints:**

- The x aesthetic should be the variable whose distribution you want to show (`price`).
- The first argument to `fct_reorder()` is the factor variable (`cut`), and the second is the numeric variable to sort by (`price`).
- Try `scale` values between 1.5 and 3 to see how the overlap changes.
- Viridis options include `"magma"`, `"plasma"`, `"inferno"`, `"viridis"`, and `"mako"`.

---

**Exercise 3: Dumbbell Chart -- Before and After**

Using the `gapminder` dataset, compare life expectancy in 1952 vs. 2007 for countries in the Americas. Create a dumbbell chart that shows both values for each country, connected by a line. Use Gonzaga red (`#C41E3A`) for 1952 and Gonzaga navy (`#002967`) for 2007.

Fill in the blanks:

```{r exercise-dumbbell-gapminder, eval=FALSE}
library(gapminder)

dumbbell_americas <- gapminder %>%
  filter(continent == "Americas", year %in% c(___, ___)) %>%
  select(country, year, lifeExp) %>%
  pivot_wider(names_from = year, values_from = lifeExp,
              names_prefix = "year_") %>%
  mutate(country = fct_reorder(country, ___))

ggplot(dumbbell_americas) +
  geom_segment(aes(x = year_1952, xend = ___, y = country, yend = ___),
               color = "#94a3b8", linewidth = 1.2) +
  geom_point(aes(x = ___, y = country), color = "#C41E3A", size = 3) +
  geom_point(aes(x = ___, y = country), color = "#002967", size = 3) +
  labs(title = "___",
       subtitle = "Red = 1952 | Navy = 2007",
       x = "Life Expectancy (years)", y = NULL) +
  theme_minimal(base_size = 12)
```

**Hints:**

- Sort countries by their 2007 life expectancy (`year_2007`) for a clean visual ordering.
- The `xend` in `geom_segment()` should be `year_2007`.
- The first `geom_point()` maps x to the 1952 column; the second maps x to the 2007 column.

---

**Exercise 4: Heatmap of mtcars Correlations**

Compute the correlation matrix for the numeric columns in `mtcars`, then reshape it into a long format suitable for `geom_tile()`. Create a heatmap of the correlations with text labels inside each tile. Use a **diverging** color scale (e.g., blue for negative, white for zero, red for positive).

Fill in the blanks:

```{r exercise-heatmap-correlations, eval=FALSE}
cor_matrix <- mtcars %>%
  select(mpg, cyl, disp, hp, drat, wt, qsec) %>%
  cor()

cor_long <- cor_matrix %>%
  as.data.frame() %>%
  rownames_to_column("var1") %>%
  pivot_longer(-var1, names_to = "var2", values_to = "correlation")

ggplot(cor_long, aes(x = var1, y = var2, fill = ___)) +
  geom_tile(color = "white") +
  geom_text(aes(label = round(___, 2)), size = 3) +
  scale_fill_gradient2(low = "___", mid = "___", high = "___", midpoint = 0) +
  labs(title = "___", fill = "Correlation") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**Hints:**

- The `fill` aesthetic and the `label` value both use the `correlation` column.
- For `scale_fill_gradient2()`, a natural choice is `low = "#002967"` (navy for negative), `mid = "white"`, and `high = "#C41E3A"` (red for positive).
- Consider adding `panel.grid = element_blank()` to remove grid lines behind the tiles.

---

**Exercise 5: Three Charts, One Dataset**

Choose a dataset and a question. Then create **three different chart types** that attempt to answer the same question. For example, if your question is "How does highway MPG vary by vehicle class?", you might create a boxplot, a violin plot, and a ridgeline plot.

For each chart, write 2--3 sentences explaining what the chart reveals and what it hides. Then write a brief paragraph (4--5 sentences) arguing which of the three is the most effective for your specific question and audience. There is no single right answer -- the point is to practice evaluating trade-offs.

Here is a starter template to get you going:

```{r exercise-three-charts-setup, eval=FALSE}
library(tidyverse)

# Choose your dataset and question:
# Dataset: ___
# Question: "How does ___ vary by ___?"

# Chart 1: ___
# (your code here)

# Chart 2: ___
# (your code here)

# Chart 3: ___
# (your code here)
```

Write your comparison paragraph below each chart in your R Markdown document.

:::
:::

## Attributions

This course material draws on and is inspired by the work of many scholars and practitioners:

- **Vivek H. Patil** -- foundational course design and materials for data visualization at Gonzaga University
- **Wilke, C.O.** -- *Fundamentals of Data Visualization* (O'Reilly, 2019), Chapters 6--12
- **Healy, K.** -- *Data Visualization: A Practical Introduction* (Princeton University Press, 2018)
- **Wickham, H.** -- *ggplot2: Elegant Graphics for Data Analysis* (Springer, 3rd edition)
- **Schwabish, J.** (2014). "An Economist's Guide to Visualizing Data." *Journal of Economic Perspectives*, 28(1), 209--234.

<div class="page-nav">
<a href="week03.html">&larr; Week 3</a>
<a href="week05.html">Week 5 &rarr;</a>
</div>
